---
title: "AlberdiLab | Carbajo et al. 2026"
subtitle: "Study in progress"
author:
  - Carlota Carbajo, Nanna Gaun, Antton Alberdi^[University of Copenhagen, antton.alberdi@sund.ku.dk], Karina Sand
date: "Last update: `r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
url: https://alberdilab.github.io/bacillus_minerals
description: |
  Bacillus minerals study
link-citations: yes
github-repo: alberdilab/bacillus_minerals
---

```{r knitr_opts, echo=FALSE}
knitr::opts_chunk$set(
    class.source = "script-source",
    class.output = "script-output",
    comment = NA)
```

# Introduction

This webbook contains all the code used for data analysis in study of gut microbiomes of newts across ponds included in a restoration plan.

## Prepare the R environment

### Environment

To reproduce all the analyses locally, clone this repository in your computer using:

```
RStudio > New Project > Version Control > Git
```

And indicating the following git repository:

> https://github.com/alberdilab/bacillus_minerals.git

Once the R project has been created, follow the instructions and code chunks shown in this webbook.

### Libraries

Install this packages if needed

```{r install_packages, eval=FALSE}
## Bioconductor packages
#https://cran.r-project.org/web/packages/BiocManager/vignettes/BiocManager.html
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

requiredPackages = c('edgeR','DESeq2','vsn','org.Hs.eg.db', "knitr")
for (p in requiredPackages) {
  if(p %in% rownames(installed.packages()) == FALSE) {BiocManager::install(p)}
}

## CRAN packages
requiredPackages = c('ggplot2','dplyr','NMF','grDevices', "hexbin", "statmod", "reshape2")
for (p in requiredPackages) {
  if(p %in% rownames(installed.packages()) == FALSE) {install.packages(p)}
}
```

The following R packages are required for the data analysis.

```{r load_libraries, warning=FALSE, comments="", message=FALSE}
# Base
library(R.utils)
library(knitr)
library(tidyverse)
library(devtools)
library(tinytable)
library(rairtable)
library(janitor)
library(broom)
library(ggtext)
library(tibble)
library(stringr)
library(dplyr)
library(DESeq2)
library(edgeR)
library(ggplot2)
library(vsn)
library(NMF)
library(org.Hs.eg.db)
library(grDevices)
library(hexbin)
library(statmod)
library(reshape2)
library(pheatmap)
library(tibble)
library(vegan)
```
